---
title: 'DT: arrange(), select(), mutate()'
author: "D. Howell"
date: "9/4/2020"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro

  * **select()** - move, drop, and rename columns
  * **mutate()** - create new columns based on existing column values
  * **arrange()** - reorder rows based on column values

Link to the data [here.](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-03-10/readme.md)

##

```{r, message=F, warning=F}
library(readr)
library(dplyr)

# reading in the tuition cost data from github using the code supplied on the README.md file
tuition_cost <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/tuition_cost.csv')
# reading in the salary potential data from github using the code supplied on the README.md file
salary_potential <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-10/salary_potential.csv')

# joining data sets
df <- inner_join(tuition_cost, salary_potential)

# initial impressions
names(df)
str(df)
summary(df)
```

##

Calculate ROI as: **[early career earnings] / [cost of tuition + room and board]**



## select()

```{r}
# selection by defining what you want to keep
df2 <- df %>%
        select(name, out_of_state_tuition, early_career_pay)
summary(df2)

# selection by defining what you want to get rid of. 
df2a <- df %>% 
          select(-(state:in_state_total), -(out_of_state_total:state_name), -(mid_career_pay:stem_percent))
summary(df2a)


```


## mutate()

```{r}

df3 <- df2 %>%
        mutate(
          roi = round((df2$early_career_pay / df2$out_of_state_tuition - 1) * 100,2)
        )

summary(df3)

```

## arrange()

```{r, message=F, warning=F}

df4 <- df3 %>%
        arrange(roi)

head(df4)


df4 <- df3 %>%
        arrange(
          desc(roi)
        )

head(df4)


df5 <- df4 %>%
        mutate(rank = 1:nrow(df4))
```

##

Let's find ORU...
```{r}
df5 %>% filter(name == 'Oral Roberts University')
```

Percentiles
```{r}

# Let's get the percentiles (the higher the better)
round((1-df5$rank[df5$name=='Oral Roberts University']/nrow(df5))*100)
round((1-df5$rank[df5$name=='University of Tulsa']/nrow(df5))*100)
round((1-df5$rank[df5$name=='Yale University']/nrow(df5))*100)
round((1-df5$rank[df5$name=='University of Central Oklahoma']/nrow(df5))*100)
```

## Further exploring
```{r}
library(ggplot2)

ggplot(df5) + 
  geom_line(aes(x = rank, y=roi))
```

##
```{r}
ggplot(df5) +
  geom_point(aes(x = rank, y = out_of_state_tuition), color = 'red') 
```

##
```{r}
ggplot(df5) + 
  geom_point(aes(x = rank, y = early_career_pay), color = 'blue')

```

## Disclaimers

  * This is out of state tuition and early career pay. I'm sure the results would be different if we chose instead to try to find lifetime earnings data about these schools. 
  * This does not factor in average scholarships given by the schools, which would probably greatly impact ROI. 
  * You might have an added constraint that you want to weed out any schools where the early career salary is less than a set amount. A program could cost $10 and the starting salary could be 12,000 which would be an incredible ROI, but that would still put you below the poverty line. 




