---
title: "EDA - Variation"
author: "D. Howell"
date: "9/14/2020"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## EDA

* Generate questions about your data.
* Search for answers by visualising, transforming, and modelling your data.
* Use what you learn to refine your questions and/or generate new questions.

## Beginning Questions

There is no universally optimal way to begin exploring your data, but in general, some basic questions that will spawn a host of new questions when answered are:

* What kind of variance is there within my variables?
* What kind of covariance is there between my variables?

## Visualizing Variance 

```{r, warning=F, message=F}
library(nycflights13)
library(dplyr)
library(ggplot2)

# create a copy of flights called df
df <- flights

# this data set has so many observations that it would take a long time to plot, so I narrow it down by finding the top 5 destinations and only keeping the observations with one of those 5 as the destination
top5_dest <- head(
  df %>%
    group_by(dest) %>%
    summarise(count = n()) %>%
    arrange(desc(count))  
  ,5
)
df_top5 <- df[df$dest %in% top5_dest$dest,]

# the PDF can be visualized with geom_freqpoly
ggplot(df_top5,aes(df_top5$dep_delay)) + 
  geom_freqpoly() 

ggplot(df_top5,aes(df_top5$dep_delay)) + 
  geom_freqpoly( binwidth = 1) 

ggplot(df_top5,aes(df_top5$dep_delay)) + 
  geom_freqpoly( binwidth = 1) + 
  coord_cartesian(xlim = c(-50, 200))

ggplot(df_top5,aes(df_top5$dep_delay)) + 
  geom_freqpoly( binwidth = 1) + 
  coord_cartesian(xlim = c(-25, 100))

ggplot(df_top5) + 
  geom_histogram(aes(df_top5$dep_delay))

ggplot(df_top5) + 
  geom_histogram(aes(df_top5$dep_delay), binwidth=1)



ggplot(df_top5,aes(df_top5$dep_delay)) + 
geom_histogram( binwidth=1)+ 
  coord_cartesian(xlim = c(-25, 100))

# what would happen if you just plotted a line of the variable instead of ordering and binning it first?
ggplot(df_top5,aes(x=1:nrow(df_top5) ,y=df_top5$dep_delay)) + 
  geom_line() 

ggplot(df_top5,aes(x=1:nrow(df_top5) ,y=df_top5$dep_delay)) + 
  geom_point() 

```

##
Another way to visualize a distribution is with a box plot

```{r, warning=F, message=F}
ggplot(df_top5,aes(y=df_top5$dep_delay)) + 
  geom_boxplot( ) + 
  coord_cartesian(ylim = c(-25, 50)) 

df_top5_filtered <- df_top5[df_top5$dep_delay<50,]
ggplot(df_top5_filtered,aes(df_top5_filtered$dep_delay)) + 
geom_histogram(binwidth=1)+ 
  coord_flip()
```

##
Visualizing the variance in a categorical variable is done with bar charts, where the y-axis represents the counts of each category within the variable. 

```{r, warning=F, message=F}
ggplot(df_top5) +  
  geom_bar(aes(x=dest))

```


## Getting the variance of other variables in flights data set

