---
title: ""
author: "D. Howell"
date: "8/26/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
if (!require(readr)) install.packages('readr')
library(readr)
library(ggplot2)

covidData <- read_csv('https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv')
```

Looks like it read in fine. Now let's filter out everything but Oklahoma.
```{r}
covidData_OK <- covidData[covidData$state == 'Oklahoma',]

ggplot(data = covidData_OK) + 
  geom_line(mapping = aes(x = date, y = cases))

```

There's some cleaning to be done here, but we'll learn how to do that later. 

Let's compare it to New York.
```{r}
covidData_NY <- covidData[covidData$state == 'New York',]

ggplot(data = covidData_NY) + 
  geom_line(mapping = aes(x = date, y = cases))
```

```{r}
covidData_yesterday <- covidData[covidData$date == as.character(Sys.Date()-1),]
ggplot(data = covidData_yesterday) + 
  geom_point(mapping = aes(x = cases, y = deaths)) + 
  geom_smooth(mapping = aes(x = cases, y = deaths))

library(dplyr)

covidData_yesterdayNoNY <- covidData_yesterday %>% filter(state != 'New York')

ggplot(data = covidData_yesterdayNoNY) + 
  geom_point(mapping = aes(x = cases, y = deaths)) + 
  geom_smooth(mapping = aes(x = cases, y = deaths))

```
There's a weird outlier way above 30k deaths. Curious what state that is. 
```{r}
covidData_yesterday[covidData_yesterday$deaths>30000,]
```
Wonder where OK is on the graph...
```{r}
covidData_yesterday[covidData_yesterday$state=='Oklahoma',]
```
So somewhere very close to the origin. Doesn't necessarily make us safer though, it is likely just because of our relatively small population. NYC has over twice as many people than OK... maybe their situation is worse in part due to the crazy population density there. 



### aesthetic mappings

I want to visualize the concentrations of covid by state, which serves to take the differences in populations out of the picture. 

```{r}
library(readr)
library(dplyr)
stPops <- read_csv('https://raw.githubusercontent.com/drehow/CSC-201/master/2020_Fall/20200821_Analytics%20with%20Excel_2/nst-est2019-alldata.csv')

stPops <- stPops[-c(1:5,nrow(stPops)),] %>% select(NAME, POPESTIMATE2019)

covidData <- left_join(covidData,stPops, by = c('state'='NAME'))

covidData$per100k <- round(covidData$cases / covidData$POPESTIMATE2019 * 10000,2)
covidData <- covidData[!is.na(covidData$per100k),]

```

```{r}
ggplot(data = covidData) + 
  geom_point(mapping = aes(x = date, y = per100k
                          #, group = state, color = state
                          ))

ggplot(data = covidData) + 
  geom_line(mapping = aes(x = date, y = per100k, group = state))
```


```{r}
 votes <- read_csv('https://raw.githubusercontent.com/drehow/CSC-201/master/2020_Fall/20200824_Beginning%20with%20R%20...%20Visualizing%20Data/1976-2016-president.csv')

library(scales)
library(tidyr)

votes <- votes %>% 
  filter(year == 2016, writein == F, (candidate == 'Trump, Donald J.' & party =='republican') | (candidate == 'Clinton, Hillary' & party=='democrat')) %>%
  select(state, party, candidatevotes) %>%
  spread(party, candidatevotes, fill=NA) 

votes <- votes %>% 
  mutate(voted = ifelse(votes$democrat > votes$republican,'democrat','republican')) %>% 
  select(state, voted)

covidData <- left_join(covidData, votes)

covidData$population <- covidData$POPESTIMATE2019
covidData$POPESTIMATE2019 <- NULL



ggplot(data = covidData) + 
  geom_line(mapping = aes(x = date, y = per100k, group = state, color = voted))

ggplot(data = covidData) + 
  geom_line(mapping = aes(x = date, y = per100k, group = state
                          , color = voted, alpha = population))

 ggplot(data = covidData) + 
  geom_line(mapping = aes(x = date, y = per100k, group = state
                          , color = voted, alpha = population
                          ))+
   labs(title="COVID Cases by State", subtitle="Data as of 8/25/2020", y="Cases per 100k People", x="Date (2020)", caption="Sources: New York Times, U.S. Census Bureau, Harvard Dataverse") 
   
ggplot(data = covidData) + 
  geom_line(mapping = aes(x = date, y = per100k, group = state
                          , color = voted, alpha = population
                          ))+
   labs(title="COVID Cases by State", subtitle="Data as of 8/25/2020", y="Cases per 100k People", x="Date (2020)", caption="Sources: New York Times, U.S. Census Bureau, Harvard Dataverse") +
    scale_alpha_continuous(labels = comma) 

ggplot(data = covidData) + 
  geom_line(mapping = aes(x = date, y = per100k, group = state
                          , color = voted, alpha = population
                          ))+
  labs(title="COVID Cases by State", subtitle="Data as of 8/25/2020", y="Cases per 100k People", x="Date (2020)", caption="Sources: New York Times, U.S. Census Bureau, Harvard Dataverse") +
  scale_alpha_continuous(labels = comma) +
  scale_x_date(date_breaks = "1 month",
                labels = date_format("%B"),
                limits = as.Date(c('2020-03-15','2020-08-23')))

ggplot(data = covidData) + 
  geom_line(mapping = aes(x = date, y = per100k, group = state
                          , color = voted, alpha = population
                          ))+
  labs(title="COVID Cases by State", subtitle="Data as of 8/25/2020", y="Cases per 100k People", x="Date (2020)", caption="Sources: New York Times, U.S. Census Bureau, Harvard Dataverse") +
  scale_alpha_continuous(labels = comma) +
  scale_x_date(date_breaks = "1 month",
                labels = date_format("%B"),
                limits = as.Date(c('2020-03-15','2020-08-23'))) +
  theme(legend.position = c(0.11, 0.67))

ggplot(data = covidData) + 
  geom_line(mapping = aes(x = date, y = per100k, group = state
                          , color = voted, alpha = population
                          ))+
  labs(title="COVID Cases by State", subtitle="Data as of 8/25/2020", y="Cases per 100k People", x="Date (2020)", caption="Sources: New York Times, U.S. Census Bureau, Harvard Dataverse") +
  scale_alpha_continuous(labels = comma) +
  scale_x_date(date_breaks = "1 month",
                labels = date_format("%B"),
                limits = as.Date(c('2020-03-15','2020-08-23'))) +
  theme_classic() +
  theme(legend.position = c(0.11, 0.67))



ggplot(data = covidData) + 
  geom_line(mapping = aes(x = date, 
                          y = per100k, 
                          group = state
                          , color = voted, 
                          alpha = population
                          )) + 
  scale_alpha_continuous(labels = comma) +
  scale_x_date(date_breaks = "1 month",
                labels = date_format("%B"),
                limits = as.Date(c('2020-03-15','2020-08-23'))) +
  scale_color_manual(values=c('#002fd9','#d90000')) +
  labs(title="COVID Cases by State", 
      subtitle="Data as of 8/25/2020", 
      y="Cases per 100k People", 
      x="Date (2020)", 
      caption="Sources: New York Times, U.S. Census Bureau, Harvard Dataverse") +
  theme_classic() +
  theme(legend.position = c(0.11, 0.67))
  

covidData_party <- covidData %>% 
  group_by(date, voted) %>% 
  summarise(per100k = mean(per100k))

ggplot(data = covidData_party) + 
  geom_line(mapping = aes(x = date, y = per100k, group = voted, color = voted))


ggplot(data = covidData[covidData$date == max(covidData$date),]) + 
  geom_boxplot(mapping = aes(x = voted, y = cases)) 
+ 
  coord_cartesian(ylim = c(0,8000))





```

### Aesthetic Mapping arguments
  * shape
  * color
  * size
  * fill
  * linetype
  * lineend
  * label
  * family (font)
  * fontface

### Sources

[Updated GGplot2 Cheat Sheet]()
[Election Data]()
[Hex Colors]()
[Used for adding labels]()
