---
title: "Functions"
author: "D. Howell"
date: "9/28/2020"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Concept
You have a data.frame object called `df`
One discrete column, with 5 categories. 
Want to perform a task only the rows of `df` where the discrete column is one of the values at a time. 

```{r, eval = FALSE}

# build the function
my_function_name <- function(argument_1){
  command_1
  command_2
  ...
}

# perform the function
my_function_name(df)

```


## Goal of Analysis
Data = mortgages

We want to make one graph for each type of mortgage showing the counts by the state in which the property is located. 

## Building our Base Data Set
We start by reading in all of our data.  

```{r, message=FALSE}
library(readr)

df1 <- read_csv('https://raw.githubusercontent.com/drehow/CSC-201/master/2020_Fall/20200928_Functions/data_1.csv')
df2 <- read_csv('https://raw.githubusercontent.com/drehow/CSC-201/master/2020_Fall/20200928_Functions/data_2.csv')
df3 <- read_csv('https://raw.githubusercontent.com/drehow/CSC-201/master/2020_Fall/20200928_Functions/data_3.csv')
df4 <- read_csv('https://raw.githubusercontent.com/drehow/CSC-201/master/2020_Fall/20200928_Functions/data_4.csv')
df5 <- read_csv('https://raw.githubusercontent.com/drehow/CSC-201/master/2020_Fall/20200928_Functions/data_5.csv')
```

rbind() for appending instead of a join

```{r}
df_all <- rbind(df1, df2, df3, df4, df5)
```

Start developing

```{r}
library(ggplot2)
library(dplyr)

# this command outputs all CONV loans
df_all %>% filter(TYPE == 'CONV')

# now we need to plot this output to get counts by state, so we need to use the geom_bar()
df_all %>% filter(TYPE == 'CONV') %>%
  ggplot() +
  geom_bar(aes(x = STATE))

# looks good. let's give it a title so that we know what type of loan this graph is for. 
df_all %>% filter(TYPE == 'CONV') %>%
  ggplot() +
  geom_bar(aes(x = STATE))  +
  labs(title = 'CONV Loans by State' ,
       y="Count", 
       x="State") 

```

## Actual Quote

> This development process would've been so much simpler if we had made a function for it at the beginning. -Every single new developer

## Rewind and use functions instead. 

Start developing . . . . 

```{r}
library(ggplot2)
library(dplyr)

# this command outputs all CONV loans
df_all %>% filter(TYPE == 'CONV')

# now we need to plot this output to get counts by state, so we need to use the geom_bar()
df_all %>% filter(TYPE == 'CONV') %>%
  ggplot() +
  geom_bar(aes(x = STATE))

# looks good. let's give it a title so that we know what type of loan this graph is for. 
df_all %>% filter(TYPE == 'CONV') %>%
  ggplot() +
  geom_bar(aes(x = STATE))  +
  labs(title = 'CONV Loans by State' ,
       y="Count", 
       x="State") 

```

Make a function to do it.
```{r}
plot_type <- function(type){
  df_all %>% filter(TYPE == type) %>%
  ggplot() +
  geom_bar(aes(x = STATE))  +
  labs(title = paste0(type,' Loans by State') ,
       y="Count", 
       x="State") 
}
```

Gotta plot 'em all
```{r}
plot_type('CONV')
plot_type('FHA')
plot_type('USDA')
plot_type('184')
plot_type('VA')
plot_type('CONV WITH MI')
plot_type('OTHER')
```




## Functions in Context

### Don't write a function when 

* you only have to perform a task once in your script
* function is performing only one or two lines of code

## Week Schedule 

* Wednesday is Conditionals 19.4
* Friday is Loops 21. 
